From Reversible Programs to \\ Univalent Universes and Back

### 2017-05-04 upi.tex
code in haskell syntax for now

First consider the following program:
\begin{verbatim}
addSub1 :: (Int, Int) :<=> (Int, Int)
addSub1 = CommuteTimes
      :.: (UnfoldN :*: Id)
      :.: Distribute
      :.: (Id :+: CommuteTimes)
      :.: Factor
      :.: (FoldN :*: Id)
\end{verbatim}
which takes a pair of natural numbers and processes them as follows depending on
whether the second number is 0 or not. If the second number is 0, the evaluation
proceeds as follows:
$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
applications of this program would increment the first number and decrement the
second number maintaining the sum at all times. When the second number reaches
0, the numbers are swapped.

As another example, consider the following expressions culminating with the
implementation of the Toffoli gate, whichi is universal for combinational
circuits.

\begin{verbatim}
inot :: Bool :<=> Bool
inot = UnfoldB :.: CommutePlus :.: FoldB

cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
cond f g = (UnfoldB :*: Id)
       :.: Distribute
       :.: ((Id :*: f) :+: (Id :*: g))
       :.: Factor
       :.: (FoldB :*: Id)

@@

toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
\end{verbatim}

@@

\begin{verbatim}
neg₁ neg₂ neg₃ neg₄ neg₅ : BOOL ⟷ BOOL
neg₁ = swap₊
neg₂ = id⟷ ◎ swap₊
neg₃ = swap₊ ◎ swap₊ ◎ swap₊
neg₄ = swap₊ ◎ id⟷
neg₅ = uniti⋆l ◎ swap⋆ ◎ (swap₊ ⊗ id⟷) ◎ swap⋆ ◎ unite⋆l
neg₆ = uniti⋆r ◎ (swap₊ {ONE} {ONE} ⊗ id⟷) ◎ unite⋆r

negEx : neg₅ ⇔ neg₁
negEx = uniti⋆l ◎ (swap⋆ ◎ ((swap₊ ⊗ id⟷) ◎ (swap⋆ ◎ unite⋆l)))
          ⇔⟨ id⇔ ⊡ assoc◎l ⟩
        uniti⋆l ◎ ((swap⋆ ◎ (swap₊ ⊗ id⟷)) ◎ (swap⋆ ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ (swapl⋆⇔ ⊡ id⇔) ⟩
        uniti⋆l ◎ (((id⟷ ⊗ swap₊) ◎ swap⋆) ◎ (swap⋆ ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ assoc◎r ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (swap⋆ ◎ (swap⋆ ◎ unite⋆l)))
          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ assoc◎l) ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ ((swap⋆ ◎ swap⋆) ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ (linv◎l ⊡ id⇔)) ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ (id⟷ ◎ unite⋆l))
          ⇔⟨ id⇔ ⊡ (id⇔ ⊡ idl◎l) ⟩
        uniti⋆l ◎ ((id⟷ ⊗ swap₊) ◎ unite⋆l)
          ⇔⟨ assoc◎l ⟩
        (uniti⋆l ◎ (id⟷ ⊗ swap₊)) ◎ unite⋆l
          ⇔⟨ unitil⋆⇔l ⊡ id⇔ ⟩
        (swap₊ ◎ uniti⋆l) ◎ unite⋆l
          ⇔⟨ assoc◎r ⟩
        swap₊ ◎ (uniti⋆l ◎ unite⋆l)
          ⇔⟨ id⇔ ⊡ linv◎l ⟩
        swap₊ ◎ id⟷
          ⇔⟨ idr◎l ⟩
        swap₊ ▤
\end{verbatim}

\begin{verbatim}
Pi... isos, recursion, finite types, Turing complete... nice blah popl 12

ESOP shows a natural 2-layer structure

Examples

\end{verbatim}

>>>>

First consider the following program:

{\footnotesize
\begin{verbatim}
addSub1 :: (Int, Int) :<=> (Int, Int)
addSub1 = CommuteTimes
      :.: (UnfoldN :*: Id)
      :.: Distribute
      :.: (Id :+: CommuteTimes)
      :.: Factor
      :.: (FoldN :*: Id)
\end{verbatim}}
%
which takes a pair of natural numbers and processes them as follows depending on
whether the second number is 0 or not. If the second number is 0, the evaluation
proceeds as follows:
$(n,0) \mapsto (0,n) \mapsto (\mathit{inj}_1(),n) \mapsto \mathit{inj}_1((),n)
\mapsto \mathit{inj}_1((),n) \mapsto (\mathit{inj}_1(),n) \mapsto (0,n)$. If the
second number is $\mathit{suc}~m$, the evaluation proceeds as follows:
$(n, \mathit{suc}~m) \mapsto (\mathit{suc}~m,n) \mapsto (\mathit{inj}_2m,n)
\mapsto \mathit{inj}_2(m,n) \mapsto \mathit{inj}_2(n,m) \mapsto
(\mathit{inj}_2n,m) \mapsto (\mathit{suc}~n,m)$. In summary, repeated
applications of this program would increment the first number and decrement the
second number maintaining the sum at all times. When the second number reaches
0, the numbers are swapped.

As another example, consider the following expressions culminating with the
implementation of the Toffoli gate, whichi is universal for combinational
circuits. We have two implementations of simple boolean negation: a direct one
and a more convoluted one. These two implementations are ``equivalent'' in the
sense to be explained below. The expression \verb|cond| is the main reversible
control flow mechanism: it allows for the conditional execution of one of two
branches but retains the choice of which branch. This expression can be used
twice to implement the Toffoli gate, which performs a conditional negation on
one bit if two other bits are both true:

{\footnotesize
\begin{verbatim}
inot :: Bool :<=> Bool
inot = CommutePlus

inot2 :: Bool :<=> Bool
inot2 = UnitTimesI
    :.: CommuteTimes
    :.: (CommutePlus :*: Id)
    :.: CommuteTimes
    :.: UnitTimesE

cond :: (a :<=> b) -> (a :<=> b) -> ((Bool, a) :<=> (Bool, b))
cond f g = Distribute
       :.: ((Id :*: f) :+: (Id :*: g))
       :.: Factor

@@

toffoli :: ((Bool,Bool),Bool) :<=> ((Bool,Bool),Bool)
toffoli = AssocTimesR :.: controlled cnot :.: AssocTimesL
\end{verbatim}}

@@

The raw syntax for combinators has, however, a significant benefit: it allows
algebraic manipulation of the programs using a second level of combinators. As
an example it is possible to show the the two implementations of boolean
negation are equivalent as follows:

{\footnotesize
\begin{verbatim}
negOpt :: inot2 :<===> inot
negOpt = (Id :>: AssocLeftEquiv)
    :..: (Id :>: (SlideCommuteTimes :>: Id))
    :..: (Id :>: AssocRightEquiv)
    :..: (Id :>: (Id :>: AssocLeftEquiv))
    :..: (Id :>: (Id :>: (InvInvEquivId :>: Id)))
    :..: (Id :>: (Id :>: IdElimL))
    :..: AssocLeftEquiv
    :..: (SlideUnitI :>: Id)
    :..: AssocRightEquiv
    :..: (Id :>: InvInvEquivId)
    :..: IdElimR
\end{verbatim}}

This program, which can be thought of as an optimization or a proof, relates the
two implementations of boolean negation via a series of elementary
optimizations of which the following are the most interesting:

{\footnotesize
\begin{verbatim}
AssocLeftEquiv    ::           (c1 :.: (c2 :.: c3)) :<===> ((c1 :.: c2) :.: c3)
SlideCommuteTimes :: (CommuteTimes :.: (c1 :*: c2)) :<===> ((c2 :*: c1) :.: CommuteTimes)
IdElimL           ::                     (Id :.: c) :<===> c
SlideUnitI        ::   (UnitTimesI :.: (c1 :*: c2)) :<===> (c2 :.: UnitTimesI)
\end{verbatim}}
